Under-approximation of reachability in multivalued asynchronous networks

### 2013-04-02 parts/intro.tex
synchronize notations of appendix B

Discrete modelling frameworks for biological networks is an active research field where formal methods have proved that they were very powerful.
Such a work started in the seventies.
It was later enriched in many directions and widely used to elucidate many biological questions.
Among these questions, a major one is to understand precisely how biological systems evolve and behave; why and how they change their usual behaviours…
This leads to questions about the accessibility (possible or inevitable) of certain states.
The ultimate goal is to discover how it could be possible to prevent from reaching some pathological states.

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised so as to be as much as possible fitting with the real systems having some observed behaviors.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Usual formal frameworks for such modelling activities are state-transition systems, Petri nets or process algebras.
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB}, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
(which would have led to a huge number of states, hopelessly far too big to be analysed).

Besides, one further objective of our work is now to be more accurate in the description of the dynamics of the studied systems.
The idea for this is to introduce timing features of a system into its model.
Indeed, we are interested in taking into account some knowledge about the relative length of some phenomena as it is a way to refute some kinds of models (parameters) non compatible with the observed dynamic behaviours.
In this paper, we are dealing with these timing properties through priorities,
that are based on the simple founding idea that the highest priority actions have to be processed before the other ones.

Until now, such a priority scheduling of the actions was not studied very much in the different formal modelling frameworks of systems biology.
Nevertheless, such an attempt has been carried out for Petri nets by F. Bause \cite{Bause97}
and the concept of priority relations among the transitions of a network has also more recently been introduced by A. K. Wagler \textit{et al.} in \cite{waw,WaglerW12} in order to allow modelling deterministic systems for biological applications.
The concept of priority is much straightforward in the approach of process algebras as it was shown by R. Cleaveland and M. Hennessy in \cite{Cleaveland199058,Cleaveland99prioritiesin} and their abstractions and equivalences were studied in \cite{Cleaveland:2007:PAP:1282576.1282847}.
It was later extended for applications in the field of system biology by M. John \textit{et al.} in \cite{jlnu2010}.

Since our formalism (Process Hitting) is inspired from the $\pi$-calculus, our work is related to such semantic ramifications of extending traditional process algebras with the concepts of priority that allow for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refer to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a two-level static priority structure, with certain ones being designated as “prioritised” and others as “unprioritised”.

Our paper is organised as follows.
The Process Hitting framework is defined in section 2;
we introduce static analysis of the PH in section 3;
section 4 illustrates the approach on an example
before the discussion and conclusion in section 5.

@@

If $A$ is a finite set,
$|A|$ is the cardinality of $A$
and $\powerset(A)$ is the power set of $A$.
If $x = (x_i)_{i \in \segm{1}{n}}$ is a sequence of elements indexed by $i \in \segm{1}{n}$,
%we denote %$|x| = (b-a)+1$ the size of this sequence and
$\indexes{x} = \segm{1}{n}$ is the set of indexes of this sequence.
We also denote by $\emptyseq$ the empty sequence.
If $A$ and $B$ are sets,
$f : A \rightarrow B$ denotes an application $f$ that maps elements of $A$ to elements of $B$.
$\lfp{x_0}{x}{x'}$ is the least fixed point greater that $x_0$ of the function $x \mapsto x'$.

>>>>

Discrete modelling frameworks for biological networks is an active research field where formal methods have proved that they were very powerful.
Such a work started in the seventies.
It was later enriched in many directions and widely used to elucidate many biological questions.
Among these questions, a major one is to understand precisely how biological systems evolve and behave; why and how they change their usual behaviours…
This leads to questions about the accessibility (possible or inevitable) of certain states.
The ultimate goal is to discover how it could be possible to prevent biological systems from reaching some pathological states.

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Usual formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB}: %, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
our motivation behind this framework was to design a model and analysis techniques that avoid to build the whole state space.
This allows us to tackle very large systems
(which would have led to a huge number of states, hopelessly too huge to be analysed).

Besides, one further objective of our work is now %to be more accurate in the description of the studied systems.
to improve the accuracy of the description of the dynamics of the studied systems.
The idea for this is to introduce timing features into its model.
Indeed, we are interested in taking into account some knowledge about the relative length of some phenomena as it is a way to refute some kinds of models (parameters) that are inconsistent with the observed dynamic behaviours.
In this paper, we are dealing with these timing properties through priorities,
that are based on the simple founding idea that the highest priority actions have to be processed before the other ones.

Until now, such a priority scheduling of the actions was not studied very much in the different formal modelling frameworks dedicated to systems biology.
Nevertheless, such an attempt has been carried out for Petri nets by F. Bause \cite{Bause97}
and the concept of priority relations among the transitions of a network has also more recently been introduced by A. K. Wagler \textit{et al.} in \cite{waw,WaglerW12} in order to allow modelling deterministic systems for biological applications.
The concept of priority is much straightforward in the approach of process algebras as it was shown by R. Cleaveland and M. Hennessy in \cite{Cleaveland199058,Cleaveland99prioritiesin} and their abstractions and equivalences were studied in \cite{Cleaveland:2007:PAP:1282576.1282847}.
It was later extended for applications in the field of system biology by M. John \textit{et al.} in \cite{jlnu2010}.

Since our formalism (Process Hitting) can be considered as a subset of CCS, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concepts of priority that allow for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refer to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a two-level static priority structure, with certain ones being designated as “prioritised” and others as “unprioritised”.

Our paper is organised as follows.
The Process Hitting framework is defined in section 2;
we introduce static analysis of the Process Hitting in section 3;
section 4 illustrates the approach on an example
before the discussion and conclusion in section 5.

@@

If $A$ is a finite set,
$|A|$ is the cardinality of $A$
and $\powerset(A)$ is the power set of $A$.
If $x = (x_i)_{i \in \segm{1}{n}}$ is a sequence of elements indexed by $i \in \segm{1}{n}$,
%we denote %$|x| = (b-a)+1$ the size of this sequence and
$\indexes{x} = \segm{1}{n}$ is the set of indexes of this sequence.
We also denote by $\emptyseq$ the empty sequence.
If $A$ and $B$ are sets,
$f : A \rightarrow B$ denotes an application $f$ that maps elements of $A$ to elements of $B$.
$\lfp{x_0}{x}{x'}$ is the least fixed point greater that $x_0$ of the function $x \mapsto x'$.
The Cartesian product is denoted $\times$.

### 2013-04-02 parts/sa.tex

\todo{Complexité de la construction du GLC}

@@

\todo{Complexité de la vérification du GLC}

>>>>

@@

Computing the graph of local causality is polynomial in the number of sorts in $\PH$ and exponential in the number of processes in one sort.
Checking the properties allowing to apply \pref{th:approxinf} is polynomial in the size of the graph.
%Therefore, if the number of processes in each sort is small, the overall method can be processed in polynomial time.
Furthermore, it is possible to compute only a subset of $\V \cap \Sol$;
in this case, the overall method turns out to be exponential in the number of solutions to each objective.
Our method can thus be considered as efficient compared to regular model-checking which is usually PSPACE-complete~\cite{Harel02}.
